---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
In keeping with the project instructions, the following code loads and provides the initial processing of the data.  

```{r}
if(!file.exists("./data")) {dir.create("./data")}

sourceFile <- "activity.zip"
if(file.exists(sourceFile)) { unzip(sourceFile, exdir="./data") }

outFile  <- "./data/activity.csv"
if (exists("DF") == FALSE) { DF <- read.csv(outFile) }
```

## What is mean total number of steps taken per day?
The distribution of the total number of steps per day is shown below.

```{r}
require(dplyr)
require(ggplot2)
require(lubridate)

ok <- complete.cases(DF)
df <- DF[ok,]
df$date <- ymd(df$date)

cdf <- aggregate(df$steps, by=list(df$date), sum)
names(cdf) <- c("Date","Steps")

g <- ggplot(cdf, aes(x = Date, y = Steps)) +
    geom_histogram(stat = "identity")
print(g)
```
The **mean** and **median** of the total number of steps per day are given below.

```{r}
summary(cdf)
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
